---
title: "netcom: Infer system functioning with empirical NETwork COMparisons"
author:
- Ryan E. Langendorf
- Debra S. Goldberg
- Matthew G. Burgess
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{netcom: Infer system functioning with empirical NETwork COMparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{dplyr}
bibliography: references.bib
csl: nature.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
#   collapse = TRUE,
  comment = "#>"
)
options(digits = 2)
```

# Abstract
Network data are ubiquitous, reflecting a burgeoning systematics paradigm across physical, natural, and social sciences. The __netcom__ package provides widely applicable tools for inferring system functioning from network data. To do this, __netcom__ compares networks to each other, across empirical classifications and theoretical models of system formation and/or functioning. This is where its name comes from: <b>net</b>work <b>com</b>parison. Patterns in the resulting state space of inter-network distances capture system functioning, risk, and underlying mechanisms.

# Installation

You can install the __netcom__ package two main ways. A release version of the package can be installed from CRAN (the Comprehensive R Archive Network): https://cran.r-project.org/package=netcom.

```{r CRAN installation, eval = FALSE}
install.packages("netcom")
```

Alternatively, the (usually) more recent development version can be installed from GitHub: https://github.com/langendorfr/netcom. This can be accomplished with the __devtools__ package. We recommend new users install the other version, from CRAN, which is has less functioning but has been more reliably tested.

```{r GitHub installation, eval = FALSE}
devtools::install_github("langendorfr/netcom")
```

Once installed simply load the package and you will have access to all of its functionality. Here we also load data manipulation and visualization packages necessary for the rest of this vignette.

```{r Setup, message = FALSE}
library(netcom)     # This library
library(igraph)     # General puprose network library
library(dplyr)      # Data manipulation
library(tibble)     # Table data format
library(expm)       # Matrix multiplication
library(ggplot2)    # Plotting
library(ggraph)     # Graph plotting extension for ggplot2
library(ggfortify)  # Statistical plotting extension for ggplot2
```

# Simulating Networks

Network simulation is a powerful way to study complex systems. By simulating networks according to different mechanisms (network generating recipes) you can study the kinds of systems that emerge. By comparing these networks to each other you can classify networks, and by comparing them to empirical network data you can quantify the probability that a particular mechanism is at play in a given real world system.

__netcom__ can simulate networks according to five common mechanisms:

1. Erdos-Renyi random (ER) @erdos1959random
2. Preferential Attachment (PA) @barabasi1999emergence
3. Duplication and Divergence (DD) @ispolatov2005duplication
4. Small-World (SW) @watts1998collective
5. Niche Model (NM) @williams2000simple

There are two ways to simulate a mechanism. (i) Grow a network adding nodes one at a time. Each new node can only use information about already existing nodes in determining which to interact with (share an edge). (ii) Take an already formed network and have a node re-choose which of the other nodes to interact with (share an edge). This is called rewiring. Notably, the sequence of rewired nodes need not be a unique ordered list of the nodes in the network. Nodes can be rewired multiple times or never. Growing networks imparts a chronological hierarchy, where no edge is bidirectional. This captures system evolution where novel entites are introduced, but is less able to describe systems of continously interacting variables. Rewired networks are more appropriate here, but can be biased by the starting network and the order in which nodes are rewired. Also, rewiring nodes enough times to reach an equilibrium can be computationally intensive.

## Mixture Networks

Note that both ways of simulating networks (growing and rewiring) involve assigning edges to one node at a time. This allows us to combine methods and mechanisms. We can simulate a network that grows, but every so often experiences rewiring, which may characterize disturbed systems. We can also simulate networks where nodes do not all act according to the same mechanism. These kinds of systems, which we call mixture networks, may be common @langendorf2020empirically.

As an example, here is a recipe for a realistically complex mixture network:

```{r Easily simulate mixture network, fig.height = 4, fig.width = 4, fig.align = "left"}
## 1. Start by creating a sequence of network evolutions. There are four components to this sequence that can be defined for each step in the network's evolution, or once which will be used for every step in the newtwork's evolution.
mechanism <- c(
    rep("ER", 7),
    rep("PA", 2),
    rep("ER", 3)
)

kind <- c(
    rep("grow", 7),
    rep("rewire", 2),
    rep("grow", 3)
)

parameter <- c(
    rep(0.3, 7),
    rep(2, 2),
    rep(0.3, 3)
)

directed <- c(
    rep(TRUE, 7),
    rep(FALSE, 2),
    rep(TRUE, 3)
)

## 2. Simulate a network according to the rules of this system evolution.
network <- make_Mixture(mechanism = mechanism, kind = kind, parameter = parameter, directed = directed)


## Finally, plot the simulated network. netcom typically stores networks as matrices,
## so to plot a network you will have to transform it into a linked list.
network %>% 
    melt() %>% 
    filter(value == 1) %>% 
    ggraph(layout = "stress") + 
        theme_void() +
        geom_node_point(size = 10,
                        color = "slateblue") +
        geom_edge_link(arrow = arrow(type = "closed",
                                     length = unit(3, "mm")),
                       start_cap = circle(4, "mm"),
                       end_cap = circle(4, "mm"))
```

This is a system that grows randomly at first, adding seven nodes that attach to other nodes according to the Erdos-Renyi (ER) random model. Note the first three elements of _kind_ are _"grow"_ signifying these are growth events. The next two events are _"rewire"_ signifying rewiring events. These do not change the size of the network (number of nodes). Only the edges between currently existing nodes are affected, as happens for the next two events where two nodes are rewired according to the Preferential Attachment (PA) model. These events include all currently existing nodes so a rewiring event can result in a node interacting with a node that was grown (added to the network) after it. Using *make_Mixture()* you can specify realistically complicated network evolutions by specifying changing governing dynamics in the sequences contained in the input parameters _mechanism_, _kind_, _parameter_, and _directed_.


# Comparing Networks

Aligning two networks is a common way to compare them to each other. This works by pairing nodes between the two networks, typically as a mathematical injection; 'one-to-one' but not necessarily 'onto'. You can visualize this as stacking one network on top of the other. A common way to do this is to minimize the disagreement in their edges. Looking at the two networks stacked together, this is the number of places where there is only one edge, indicating that one of the networks claims there is an interaction between the two nodes connected by the edge and the other thinks those two nodes describe variables that do not interact.

Network alignment offers useful insights at both the level of the entire network and the level of individual nodes. Paired nodes suggest variables that may serve similar functional roles in their respective systems, and may even respond similarly to disturbances. This kind of information sharing is particularly helpful in systems that are difficult to study but have closely related systems that are more common, for example a species that is endangered in one country but common in another. Zooming out to the level of entire networks, network alignment offers a way to map state spaces of systemic properties, structures, and functions. 

```{r Align networks, fig.height = 4, fig.width = 4, fig.align = "left"}
## Create two sets of 5 networks each, all with 10 nodes. One of these sets is grown
## according to the Small-World (SW) random mechanism. The other is also grown in this
## manner, but with one randomly rewired nodes for every 2 nodes grown, mimicking a 
## disturbed system.
num_networks <- 10

## First set of undisturbed networks.
## You can omit the '_grow' suffix which is the default, as long as every node is the be grown.
networks_undisturbed <- list()
for (net in 1:num_networks) {
    networks_undisturbed[[net]] = make_Mixture(
        mechanism = rep("DD", 20), 
        parameter = 0.15, 
        directed = TRUE
    )
}

## Second set of disturbed networks
networks_disturbed <- list()
for (net in 1:num_networks) {
    networks_disturbed[[net]] = make_Mixture(
        mechanism = rep(c("DD", "DD", "DD", "ER", "ER", "ER", "DD", "DD"), 4),
        parameter = rep(c(0.3, 0.3, 0.3, 0.76, 0.76, 0.76, 0.1, 0.1), 4),
        kind = rep(c("grow", "grow", "grow", "rewire", "rewire", "rewire", "grow", "grow"), 4),
        directed = TRUE
    )
}

## Pairwise compare all of the networks to each other 
networks <- c(networks_undisturbed, networks_disturbed)
comparisons <- matrix(NA, nrow = length(networks), ncol = length(networks))
for (row in 1:length(networks)) {
    for (col in 1:length(networks)) {
        
        ## Take advantage of the symmetry of this alignment algorithm
        if (col > row) {
            comparisons[row, col] = align(networks[[row]], networks[[col]])$score
            comparisons[col, row] = comparisons[row, col]
        }

    }
}
## Self-alignment always returns a zero distance so these can be added manually
diag(comparisons) = 0

## PCA of comparisons
stats::prcomp(comparisons) %>%
    autoplot(
        data = tibble(
            Kind = c(
                rep("Duplication & Divergence", num_networks),
                rep("Disturbed Duplication & Divergence", num_networks)
            )
        ), 
        colour = "Kind",
        size = 5
    ) + 
    theme_bw() +
    theme(
        legend.position = "bottom",
        legend.title = element_blank()
    )
```

Aligning networks is not the only way to compare them. For example, the average degree (number of interactions) in a network can be compared. This does not tell you which nodes are most like each other between the two networks, but does allow you to make network-level inferences. It also more naturally allows for ensemble methods, where you compare networks as a weighted average of the differences across multiple characterizations.

# Network Classification
We can combine network simulation and comparisons to test a hypothesized mechanism against network data. Here the p-values are for the null hypothesis that the network was generated according to the particular mechanism being tested. Small p-values indicate the mechanism likely does not govern the system described by the network. Note that non-significant p-values do not prove the network was generated by that mechanism, only that it appears no more dissimilar to networks that are generated by that mechanism than they are to each other.

```{r Classify networks, fig.height = 4, fig.width = 4, fig.align = "left"}
## Classification of an undisturbed Small-World network made above
classify(
    network = networks_undisturbed[[1]], 
    processes = c("SW", "DD", "NM", "PA", "ER", "DD"),
    directed = TRUE,
    mechanism_kind = "grow"
)

## Classification of a disturbed Small-World network made above
classify(
    network = networks_disturbed[[1]], 
    processes = c("SW", "DD", "NM", "PA", "ER", "DD"),
    directed = TRUE,
    mechanism_kind = "grow"
)
```

Pure mechanisms are readily inferred by comparing a network with many networks simualted from hypothesized mechanisms. Mixture mechanisms are not. This is because different mixtures can produce similar network structures. Fortunately, we can still functionally classify network data describing realistically complex systems governed by mixtures of mechanisms @langendorf2020empirically.

# References