---
title: "netcom: Infer system functioning with empirical NETwork COMparisons"
author:
- Ryan E. Langendorf
- Debra S. Goldberg
- Matthew G. Burgess
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{netcom: Infer system functioning with empirical NETwork COMparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
bibliography: references.bib
csl: nature.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
#   collapse = TRUE,
  comment = "#>"
)
options(digits = 2)
```

# Abstract
Network data are ubiquitous, reflecting a burgeoning systematics paradigm across physical, natural, and social sciences. Means of inferring the functioning of the systems these networks describe has lagged behind. The **netcom** package provides generally applicable tools for inferring system functioning from network data. Instead of testing models of system formation and/or functioning with network data, **netcom** compares the networks to each other. Patterns in the resulting state space of inter-network distances capture system functioning, risk, and underlying mechanisms.

# Installation

You can install the **netcom** package two main ways. A release version of the package can from CRAN (the Comprehensive R Archive Network): https://cran.r-project.org/package=netcom.

```{r CRAN installation, eval = FALSE}
install.packages("netcom")
```

Alternatively, the most recent version can be installed from GitHub: https://github.com/langendorfr/netcom. This can be accomplished with the **devtools** package.

```{r GitHub installation, eval = FALSE}
devtools::install_github("langendorfr/netcom")
```

Once installed simply load the package and you will have access to all of its functionality.

```{r Setup}
library(netcom)
```

# Simulating Networks

Network simulation is a powerful way to study complex systems. By simulating networks according to different mechanisms (network generating recipes) you can study the kinds of systems that emerge. By comparing these networks to each other you can classify networks, and by comparing them to empirical network data you can quantify the probability that a particular mechanism is at play in a given real world system.

__netcom__ can simulate networks according to five common mechanisms:

1. Erdos-Renyi random (ER) @erdos1959random
2. Preferential Attachment (PA) @barabasi1999emergence
3. Duplication and Mutation (DM) @ispolatov2005duplication
4. Small-World (SW) @watts1998collective
5. Niche Model (NM) @williams2000simple

There are two ways to simulate a mechanism. (i) Grow a network adding nodes one at a time. Each new node can only use information about already existing nodes in determining which to interact with (share an edge). (ii) Take an already formed network and have a node re-choose which of the other nodes to interact with (share an edge). This is called rewiring. Notably, the sequence of rewired nodes need not be a unique ordered list of the nodes in the network. Nodes can be rewired multiple times or never. Growing networks imparts a chronological hierarchy, where no edge is bidirectional. This captures system evolution where novel entites are introduced, but is less able to describe systems of continously interacting variables. Rewired networks are more appropriate here, but can be biased by the starting network and order in which nodes are rewired. Also, rewiring nodes enough times to reach an equilibrium is computationally intensive.

## Mixture Networks

Note that both ways of simulating networks (growing and rewiring) involve assigning edges to one node at a time. This allows us to combine methods and mechanisms. We can simulate a network that grows, but every so often experiences rewiring, which may characterize disturbed systems. We can also simulate networks where nodes do not all act according to the same mechanism. These kinds of systems, which we call mixture networks, may be common @langendorf2020empirically.

As an example, here is a recipe for a realistically complex mixture network:

```{r Simulate mixture network}
## 1. Start by creating a network of the desired final size with only two variables that
## interact with each other.
network_size <- 10
network <- matrix(0, nrow = network_size, ncol = network_size)
network[1, 0] = 1
network[0, 1] = 1

## 2. Grow five new nodes. All ten follow the Erdos-Renyi random mechanism, but with a 
## decreasing probability of interaction, going from 1 to 0.1.
p_interaction <- 1
for (node in 3:7) {
    network = netcom::grow_ER(network, x = node, p = p_interaction)
    p_interaction = p_interaction - 0.1
}

## 3. Rewire three of the nodes using preferential attachment a power of 2.
rewire_ids <- sample(1:7, 3, replace = FALSE)
for (node in rewire_ids) {
    network_exists <- network[1:7, 1:7]
    network_exists <- netcom::stir_PA(network_exists, x = node, power = 2)
    network[1:7, 1:7] <- network_exists
}

## 4. Grow 3 new nodes (10 total) using the Erdos-Renyi random mechanism again. All eight 
## have a probability of attachment of 0.5.
p_interaction <- 0.5
for (node in 8:10) {
    network = netcom::grow_ER(network, x = node, p = p_interaction)
}

print(network)
```

This models a community that grows randomly, but with a decreasing chance of new members interacting. Several choose to rethink their interactions, preferring to interact with members that have lots of interactions. Subsequent growth then involves a constant higher level of interactions with new members.

Initializing an empty matrix, and having to specify which subset of the matrix to use in a rewiring event, may seem cumbersome but they allow you to simulate a greater diversity of networks. For example, you can grow nodes anachronistic to their location in the network matrix, and explore rewiring in the context of metacommunities, neither of which would be possible if the input network was itself a fully simulated network.

# Comparing Networks

Aligning two networks is a common way to compare them to each other. This works by pairing nodes between the two networks, typically as a mathematical injection; 'one-to-one' but not necessarily 'onto'. You can visualize this by imagining stacking one network on top of the other. A common way to do this is to minimize the disagreement in their edges. Looking at the two networks stacked together, this is the number of places where there is only one edge, indicating that one of the networks claims there is an interaction between the two nodes connected by the edge and the other thinks those two nodes describe variables that do not interact.

Network alignment offers useful insights at both the level of the entire network and the level of individual nodes. Paired nodes suggest variables that may serve similar functional roles in their respective systems, and may even respond similarly to disturbances. This kind of information sharing is particularly helpful in systems that are difficult to study but have closely related systems that are more common, for example a species that is endangered in one country but common in another. Zooming out to the level of entire networks, alignments offer a way to map state spaces of systemic properties, structures, and functions. 

```{r Align networks}
```

Aligning networks is not the only way to compare them. For example, the average degree (number of interactions) in a network can be compared. This does not tell you which nodes are most like each other between the two networks, but does allow you to make network-level inferences. It also more naturally allows for ensemble methods, where you compare networks as a weighted average of the differences across multiple characterizations.

# Network Classification
We can combine network simulation and comparisons to test a hypothesized mechanism against network data.

```{r Classify networks}
```


# Acknowledgements

**netcom** is the product of years of collaborative research. We thank: Dan Doak, ...

# References